Heavy rainfall in Paraguay during the 2015-2016 austral summer: causes and sub-seasonal-to-seasonal predictive skill

### 2018-03-20 reports/writeup/paper.tex
James edits of Angel comments

\abstract{
	During the austral summer 2015-16, severe flooding displaced over \num{170000} people on the Paraguay River system in Paraguay, Argentina, and Southern Brazil.
	These floods were driven by repeated heavy rainfall events in the Lower Paraguay River Basin.
	Alternating sequences of enhanced moisture inflow from the South American Low-Level Jet and local convergence associated with baroclinic systems were conducive to mesoscale convective activity and enhanced precipitation.
	These circulation patterns were favored by cross-timescale interactions of a very strong El Ni\~{n}o event, an unusually persistent Madden-Julian Oscillation in phases four and five, and the presence of a dipole SST anomaly in the central southern Atlantic Ocean.
	The simultaneous use of seasonal and sub-seasonal heavy rainfall predictions could have provided decision makers useful information about the start of these flooding events from at least two-to-four weeks in advance.
	Probabilistic seasonal forecasts available at the beginning of November successfully indicated heightened probability of heavy rainfall (\nth{90} percentile) over southern Paraguay and Brazil for December-February.
	Raw sub-seasonal forecasts of heavy rainfall exhibited limited skill at lead times beyond the first two predicted weeks, but a Model Output Statistics approach involving principal component regression substantially improved the spatial distribution of skill for Week 3 relative to other methods tested including extended logistic regressions.
	A continuous monitoring of climate drivers impacting rainfall in the region, and the use of bias-corrected heavy precipitation seasonal and sub-seasonal forecasts, may help improve flood preparedness for the austral summer season in this part of the world.
}

@@

During the austral summer of 2015-16, repeated heavy rainfall events led to severe flooding in the \gls{lprb}, displacing approximately \num{170000} people \citep{Brakenridge2016} and causing tremendous damage to property and infrastructure \citep{MinisteriodeObrasPublicasyComunicacion2016}.
Because population in South America tends to concentrate along coasts and rivers (supplemental figure S1\footnote{check}), flooding in the \gls{lprb} directly affects not only much of the population of Paraguay, but also of populations in Argentina and Uruguay who lie along the Paran\'{a} and la Plata rivers, into which the Paraguay River drains.
Heavy rainfall and flooding in the \gls{lprb} also has important implications for hydropower generation, for agriculture, and for regional water resource management.
The aim of this paper is to diagnose the drivers of the November-February (NDJF) 2015-16 rainfall and flooding events, and to assess the skill of the associated subseasonal-to-seasonal predictions.

The rainfall climatology of the \gls{lprb} varies strongly by season, with extratropical characteristics in the winter and monsoonal characteristics in the summer.
In the warm season (NDJF), which is the focus of this study, circulation is dominated by the upper-tropospheric Bolivian High, the lower-level subtropical highs, the Chaco Low over northern Argentina, the \gls{sacz}, and the \gls{sallj} \citep{Grimm2009,Marengo2012}.
Rainfall peaks around \SI{5}{\milli\meter\per\day} during the warm months (October-May) and reaches a minimum near \SI{2}{\milli\meter\per\day} in July and August.
However, the flat topography limits the river's ability to carry the summer runoff, causing seasonal inundation of the Pantanal and distributing the river discharge in time \citep{Bravo2011,Barros2004}.
Thus, the streamflow maxima typically occurs in phase with precipitation upstream of the Pantanal, while downstream of the Pantanal - which we define as the \acrlong{lprb}; see \cref{fig:study-area} - the annual peak typically occurs between April and July.

During the warm season, a large fraction of rainfall, and nearly all heavy rainfall, in the \gls{lprb} is associated with mesoscale convection \citep{Velasco1987}.
Previous studies of organized convection and precipitation across subtropical continental South America have found close correspondence with the exit region of the low-level jets \citep{Saulo2007,Salio2007,Marengo2004,Velasco1987}, which is influenced in both summer and winter by mid-latitude baroclinic wave trains that interact with the Andes topography to generate orographically bound cyclones and northerly low-level flow \citep{Campetella2002,Seluchi2006,Boers2013,Boers2014}.
The strength and direction of this moisture transport varies substantially between events; \gls{sallj} exit regions range from central Argentina \citep[``Chaco Jet Events'';][]{Salio2002} to Paraguay and southeastern Brazil \citep[``No-Chaco Jet Events'';][]{Vera2006}.

At sub-seasonal timescales, convection in the \gls{lprb} is modulated by a variety of drivers, notably including the \gls{sacz} and the \gls{mjo}.
During \gls{sacz} conditions, strong low-level convergence is observed over the Amazon basin with low-level divergence over southwestern Brazil, northern Argentina and Paraguay \citep{Herdies2002,Carvalho2010}; the opposite is true for so-called No-\gls{sacz} conditions.
\Gls{sacz} occurrence is related to westerly wind regimes over South East South America, as well as ``active'' and ``break'' periods of the South American Monsoon System \citep{Marengo2004}.
The \gls{mjo} has been associated with the South American ``seesaw'' pattern \citep{Nogues-Paegle1997,Paegle2000,Liebmann2004}, and has been identified as a source of rainfall predictability for the region \citep[e.g.][]{Munoz2015}.

At seasonal timescales, \gls{enso} is the dominant driver of convection variability in the \gls{lprb}.
During El Ni\~no years, a low-level anticyclonic anomaly over central Brazil enhances occurrence of the low-level jet, favoring the development of mesoscale convective systems \citep{Velasco1987}.
The intensity and precise extent of this anomaly is relevant for the precise impact of \gls{enso} events.
The region also exhibits substantial variability between seasons of rainfall during El Ni\~no years, including a reversal of rainfall anomalies between November of that year and January of the following one, influenced by land-surface interactions \citep{Grimm2009,Grimm2003}.
Even beyond El Ni\~no years, regional land-surface feedbacks can cause regions that exhibit wet anomalies in the spring to experience more summer precipitation on average \citep{Grimm2007}.
Similarly, mid-latitude dynamics influence low-level wind anomalies on many time scales, though this relationship is complicated due to coupled tropical-extratropical interactions \citep{Carvalho2004,Jones2002}.
To address these potential interactions, a cross-timescale approach based on synoptic circulation types is employed here to diagnose the causes of the rainfall events; this method has been used in previous work for southeastern South America \citep{Munoz2015,Munoz2016} and other regions \citep{Moron2015}.

We begin by describing our data sources in \cref{sec:data} and our methods in \cref{sec:methods}.
In \cref{sec:diagnostics} we start our diagnosis highlighting the observed flooding and contextualizing it within a long river stage time series; then we use composites and a weather typing analysis to diagnose the circulation patterns associated with the heavy rainfall during NDJF 2015-16. We turn in \cref{sec:fcsts} to the question of whether the observed rainfall was successfully predicted by available models.
To carry out this analysis we study both forecasts targeting the entire series for a limited area, and also forecasts targeting a large spatial area for only the first week of December, when the most important flooding events started.
We also explore the impact on forecasts of several bias-correction schemes.
In \cref{sec:discussion} we discuss limitations and potential implications of our findings and potential future work, and we present our concluding remarks in \cref{sec:summary}.

@@

The analysis presented involves both observations and model forecasts.

@@

Rainfall data are taken from the CPC Unified Gauge-based Analysis of Global Daily Precipitation dataset \citep{Chen2008}.
Spatial resolution is \SI{0.5}{\degree}, and temporal resolution is daily.
We define ``heavy'' rainfall events to be exceedances of the \nth{90} percentile; while the value is different for each grid cell, the \nth{90} percentile of area-averaged rainfall over the \gls{lprb} is approximately \SI{15}{\milli\meter\per\day}.

Atmospheric circulations are diagnosed using daily data from the NCAR-NCEP Reanalysis II dataset \citep{Kanamitsu2002}.
Spatial resolution is \SI{2.5}{\degree}.
Because the end-of-day time for the rainfall data is 12:00 GMT over most of South America \citep{Chen2008}, we use six-hour reanalysis data, and shift by twelve hours before re-sampling to the daily time step.
This ensures that the time steps in the reanalysis and ƒrainfall data sets are the same, but means that a day is defined as beginning at 12:00 GMT.
Since most summer rainfall in this region occurs overnight \citep{Salio2007,Vera2006}, this end-of-day time (which translates to approximately 8:00 AM locally depending on the exact time zone) tends to separate distinct events.
The primary atmospheric variable  used was the \SI{850}{\hecto\pascal} streamfunction, calculated directly from the wind field as described in \cref{sec:methods}.
The streamfunction is preferable to, for example, the geopotential height $\Phi$ because $\Phi$ has weak gradients near the equator, making it difficult to visualize circulations that span from the tropics to the extratropics.
The \SI{850}{\hecto\pascal} height level was used because it is representative of \gls{sallj} activity and moisture transport in this region \citep{Marengo2004,Salio2007}.

@@

This study also makes use of some climate indices.
Data on \gls{enso}, specifically the NI\~NO3.4 index, came from a statistical-dynamical interpolation \citep{Kaplan1998}, which is constrained by relatively high-quality observations during the study period.
Data on the \gls{mjo}   came from the Australian Bureau of Meteorology \citep{Wheeler2004}.

@@

The seasonal predictions used  are known as ``flexible format'' forecasts, provided by the \gls{iri}.
These forecasts use a multi-model ensemble approach, with bias-corrected retrospective probabilistic forecasts produced using a total of (a) 144 members forced by evolving sea-surface temperatures, and (b) 68 members forced by persisted sea-surface temperatures; for details see \citet{Barnston2010}.
Flexible format means that the user of these forecasts can arbitrarily choose  particular thresholds (percentiles) to compute the  probability of exceedance (or non-exceedance) from the complete  probability density function of the climatological distribution,  rather than using the more common tercile categories.
The DJF 2015-2016 forecasts analyzed were produced in  November 2015.
Due to the short sample of flexible format forecasts available (only for 2012-2016) at the time of writing  this paper, no verification was performed for these seasonal predictions.
These forecasts are provided at a horizontal  resolution of \SI{2.5}{\degree}.
This product provides the probability of exceedance (or non-exceedance) of a user-specified percentile of the climatological distribution.
The DJF 2015-2016 forecasts analyzed were produced in November 2015.

@@

Several types of analyses are used to diagnose the causes of the heavy rainfall events, and to bias-correct and verify the forecasts.
Computation was performed in the python environment using stable open source packages \citep{Hoyer2017,vanderWalt2011,McKinney2010,Hunter2007}.
All codes to reproduce or modify this analysis are available at \url{https://github.com/jdossgollin/PYFloods}.

@@

The $K$-means algorithm is guaranteed to converge to a local minimum of inter-cluster variance; to select the best partition, 500 simulations were created using the implementation in Python's \texttt{scikit-learn} package \citep{Pedregosa2012} and then the classifiability index of \citet{Michelangeli1995} was computed between each partition and the 499 others.
The partition whose classifiability index, averaged for all 499 pairwise comparisons, was the highest was selected.
Calculation of the classifiability index for several values of $K$ (supplemental figure S3)\footnote{CHECK} suggests that states with $K=5,6,\ldots,8$ are all reasonable.
We chose the solution $K=6$ because the clusters identified are qualitatively similar to those determined over southeastern South America \citep{Munoz2015,Munoz2016} and have an intuitive physical meaning, discussed further in the following sections.
We refer to the resulting clusters as \glspl{wt}.
From a physical point of view, the $K$-means algorithm helps identify typical atmospheric circulation patterns in the \gls{eof}-filtered field via clustering of days with similar streamfunction configurations.
These clusters can also be understood as proxies  of the available states of the system , or the most frequently visited trajectories in the phase space of the physical system \citep{Munoz2015,Munoz2016a,Munoz2017}.

@@

Logistic regression models the conditional probability of binary events, and it has been widely used in \gls{mos} approaches for binary outcomes.
Nonetheless, when using the logistic regression to address multiple thresholds via independent fits, the predicted probabilities are, in general, not mutually consistent \citep{Messner2014}.
The \gls{xlr} was designed to address this shortcoming \citep{Wilks2009}, via the consideration of a transformation of the thresholds of interest as an additional predictor variable.
The \gls{hxlr}, a generalization of the \gls{xlr}, was proposed to appropriately use the ensemble spread as predictor for the dispersion of the predictive distribution \citep{Messner2014}.

@@

To avoid artificial skill when building the statistical models, we employ a cross-validation approach with a 5-year window.
In this framework, five continuous years are left out of the record, the regression coefficients are computed with the remaining of the time series, and the resulting model is validated comparing the prediction for the third year left out (middle of the window) against observations.
The 5-year-long window is redefined a year at a time, moving from the beginning of the record to its end.

@@

\Cref{fig:streamflow} shows the streamflow time series at several gauges on the Paraguay River during NDJF 2015-16 in the context of their seasonality and decadal variability.
During November and December 2015, the river rose rapidly at Concepci\'on, Asunci\'on, and Pilar though not at Bah\'ia Negra.
As discussed in \citet{Bravo2011,Barros2004}, the location of the Bah\'ia Negra gauge (see \cref{fig:study-area}) in the Pantanal region means that it responds very slowly to rainfall input.
The three downstream gauges, because they are located in the \gls{lprb}, respond to the rainfall forcing with a slow but steady rise.
Despite several very heavy storms, the streamflow record at Asunci\'on and Pilar (which are downstream of Concepci\'on) indicates relatively little response to individual storms.
Because the region is so flat (see topographic data in \cref{fig:study-area}), river levels at a particular point may be affected not only by rain in the catchment corresponding to that point, but also by elevated river levels downstream which reduce the pressure gradient available to drive flow.

@@

\Cref{fig:lagged-rain} shows time-lagged anomalies up to and after heavy rainfall dates (when area-averaged daily rainfall in the \gls{lprb} exceeds its NDJF \nth{90} percentile) and is consistent with previous analysis of heavy rainfall and intense convection in this region \citep{Liebmann2004,Marengo2004,Salio2007,Marwan2015}.
At $t=-2$ a mid-latitude baroclinic system approaches the South American continent, intensifying and moving to the East at $t=-1,0,1$ days.
This system interacts with the sub-tropical low and the Andes Mountains to produce an anticyclonic anomaly over Brazil.
Along the cold front associated with this system, a low-level northerly jet advects  heat and moisture to the region.
As the system progresses, the jet below \SI{20}{\degree S} transitions from predominantly meridional flow (``Chaco Jet''; $t=-1$ to predominantly zonal flow (``No-Chaco Jet'').
The pattern resembles composites identified using one standard deviation exceedances of rainfall at \SI{60}{\degree W}, \SI{30}{\degree S} \citep{Liebmann2004} and analysis for the \nth{95} or 99th percentiles of daily rainfall (not shown) yield similar results, implying that the synoptic mechanism for the most heavy events is not fundamentally distinct from the mechanism for moderate-intensity events.
This mean field, like all composites, masks between-event variation but exploration of individual events (not shown) indicates that the core features identified are generally present.

@@

\Gls{wt} 1 describes a \gls{sallj} event in which the strongest wind penetrates southward of \SI{25}{\degree S}, leading to heavy rainfall over NE Argentina and Uruguay; this has been called a ``Chaco Jet'' event \citep{Salio2002}.
\Gls{wt} 4 is also associated with a \gls{sallj} event, but the wind turns to the East northward of \SI{25}{\degree S}, leading to heavy rainfall over Eastern Paraguay and SW Brazil; this has been called a ``No-Chaco'' Jet event \citep{Vera2006}.
Supplemental Table S2\footnote{Check} shows the centroids of each cluster, in the 4-dimensional phase space of the leading \glspl{eof} of \SI{850}{\hecto\pascal} streamfunction.

@@

We next use monthly-mean circulation anomalies (spatial patterns) and weather type sequences (time patterns) to understand the specific events of NDJF 2015-16.

@@

As indicated in \cref{fig:rain-wt}, the most heavy rainfall occurred during \glspl{wt} 1 and 4.
During NDJF 2015-16 \glspl{wt} (Chaco and No-Chaco jet extensions, respectively),  occurred more frequently than their climatology (supplemental table S1)\footnote{check}; \gls{wt} 2 also occurred more frequently than its climatology, largely due to a long sequence in February 2016.
In mid-January 2016, during a sequence of persistent low rainfall, \gls{wt} 3 featured persistently, leading to heavy rainfall over central Brazil (not shown) and negative rainfall anomalies over the \gls{lprb}.
Thus while the intensity and persistence of heavy rainfall was atypical, the causal mechanism of the heavy rainfall observed during this season was consistent with climatology.

Inspection of \cref{fig:rain-wt} also suggests that at time scales of days to weeks, particular sequences of weather types tend to recur, and are associated with repeated rainfall storms.
The rainfall in late November 2015 occurred during \gls{wt} 1 and 4 days, with intermittent (and dry) \gls{wt} 5 occurrences.
During mid to late December 2015, \glspl{wt} 1 and 5 produced rainfall but the intermittent dry days were \gls{wt} 2.
During mid to late January 2016 \gls{wt} 3 led to persistent low rainfall.

@@

Heavy rainfall over the region was forecast for the DJF 2015-2016 season since at least November 2015 (see \cref{fig:seas-prob-fcst}).
Relative odds as high as 7:1 are visible over southern Paraguay and Brazil, and northern Uruguay and Argentina, generally in agreement with observations.
The model predicted only very weakly increased odds of heavy rainfall in the Pantanal region (directly north of the \gls{lprb}) and in northern Argentina at $\approx \SI{65}{\degree W}$, and missed the heavy precipitation along most of the northeastern border of Paraguay.
However, the regionally elevated forecast of heavy rainfall could have been used for disaster preparedness at least one month in advance.

@@

\Cref{fig:chiclet} uses a Chiclet diagram \citep{Carbin2016} to visualize, as a function of lead time, the time evolution of the uncorrected, ensemble-mean rainfall anomaly forecast, spatially averaged over the \gls{lprb}.
At times greater than about two weeks, the ensemble-mean forecast is slightly wetter than climatology.
At weather timescales (less than one week), the ensemble-mean successfully predicts the timing and amplitude of the area-averaged rainfall.
At timescales of one to three weeks, the model successfully forecast the strongest breaks and pauses in the rainfall, such as the heavy rainfall during December 2015 and the dry period during mid-January 2016.

>>>>

\abstract{
	During the austral summer 2015-16, severe flooding displaced over \num{170000} people on the Paraguay River system in Paraguay, Argentina, and Southern Brazil.
	These floods were driven by repeated heavy rainfall events in the Lower Paraguay River Basin.
	Alternating sequences of enhanced moisture inflow from the South American Low-Level Jet and local convergence associated with baroclinic systems were conducive to mesoscale convective activity and enhanced precipitation.
	These circulation patterns were favored by cross-timescale interactions of a very strong El Ni\~{n}o event, an unusually persistent Madden-Julian Oscillation in phases four and five, and the presence of a dipole SST anomaly in the central southern Atlantic Ocean.
	The simultaneous use of seasonal and sub-seasonal heavy rainfall predictions could have provided decision makers useful information about the start of these flooding events from at least two-to-four weeks in advance.
	Probabilistic seasonal forecasts available at the beginning of November successfully indicated heightened probability of heavy rainfall (\nth{90} percentile) over southern Paraguay and Brazil for December-February.
	Raw sub-seasonal forecasts of heavy rainfall exhibited limited skill at lead times beyond the first two predicted weeks, but a Model Output Statistics approach involving principal component regression substantially improved the spatial distribution of skill for Week 3 relative to other methods tested including extended logistic regressions.
	A continuous monitoring of climate drivers impacting rainfall in the region, and the use of bias-corrected heavy precipitation seasonal and sub-seasonal forecasts, may help improve flood preparedness in this and other regions.
}

@@

During the austral summer of 2015-16, repeated heavy rainfall events led to severe flooding in the \gls{lprb}, displacing approximately \num{170000} people \citep{Brakenridge2016} and causing tremendous damage to property and infrastructure \citep{MinisteriodeObrasPublicasyComunicacion2016}.
Because population in South America tends to concentrate along coasts and rivers (supplemental figure S1), flooding in the \gls{lprb} directly affects not only much of the population of Paraguay, but also of populations in Argentina and Uruguay who lie along the Paran\'{a} and la Plata rivers, into which the Paraguay River drains.
Heavy rainfall and flooding in the \gls{lprb} also has important implications for hydropower generation, for agriculture, and for regional water resource management.
The aim of this paper is to diagnose the drivers of the November-February (NDJF) 2015-16 rainfall and flooding events, and to assess the skill of the associated subseasonal-to-seasonal predictions.

The rainfall climatology of the \gls{lprb} varies strongly by season, with extratropical characteristics in the winter and monsoonal characteristics in the summer.
In the warm season (NDJF), which is the focus of this study, circulation is dominated by the upper-tropospheric Bolivian High, the lower-level subtropical highs, the Chaco Low over northern Argentina, the \gls{sacz}, and the \gls{sallj} \citep{Grimm2009,Marengo2012}.
Rainfall peaks around \SI{5}{\milli\meter\per\day} during the warm months (October-May) and reaches a minimum near \SI{2}{\milli\meter\per\day} in July and August.
However, the flat topography limits the river's ability to carry the summer runoff, causing seasonal inundation of the Pantanal and distributing the river discharge in time \citep{Bravo2011,Barros2004}.
Thus, the streamflow maxima typically occurs in phase with precipitation upstream of the Pantanal, while downstream of the Pantanal - an area which we define in \cref{fig:study-area} as the \acrlong{lprb} - the annual peak typically occurs between April and July.

During the warm season, a large fraction of rainfall, and nearly all heavy rainfall, in the \gls{lprb} is associated with mesoscale convection \citep{Velasco1987}.
Previous studies of organized convection and precipitation across subtropical continental South America have found close correspondence with the exit region of the low-level jets \citep{Velasco1987,Marengo2004,Saulo2007,Salio2007}, which is influenced in both summer and winter by mid-latitude baroclinic wave trains that interact with the Andes topography to generate orographically bound cyclones and northerly low-level flow \citep{Campetella2002,Seluchi2006,Boers2013,Boers2014}.
The strength and direction of this moisture transport varies substantially between events, and \gls{sallj} exit regions range from central Argentina \citep[``Chaco Jet Events'';][]{Salio2002} to Paraguay and southeastern Brazil \citep[``No-Chaco Jet Events'';][]{Vera2006}.

At sub-seasonal timescales, heavy rainfall and convection in the \gls{lprb} is modulated by a variety of drivers, notably including the \gls{sacz} and the \gls{mjo}.
During \gls{sacz} conditions, strong low-level convergence is observed over the Amazon basin with low-level divergence over southwestern Brazil, northern Argentina and Paraguay \citep{Herdies2002,Carvalho2010}; the opposite is true for so-called No-\gls{sacz} conditions.
\Gls{sacz} occurrence is related to westerly wind regimes over South East South America, as well as ``active'' and ``break'' periods of the South American Monsoon System \citep{Marengo2004}.
The \gls{mjo} has been associated with the South American ``seesaw'' pattern \citep{Nogues-Paegle1997,Paegle2000,Liebmann2004}, and has been identified as a source of rainfall predictability for the region \citep[e.g.][]{Munoz2015}.

At seasonal timescales, \gls{enso} is the dominant driver of convection variability in the \gls{lprb}.
During El Ni\~no years, a low-level anticyclonic anomaly over central Brazil enhances occurrence of the low-level jet, favoring the development of mesoscale convective systems \citep{Velasco1987}.
The intensity and precise extent of this anomaly is relevant for the precise impact of \gls{enso} events.
The region also exhibits substantial variability between seasons of rainfall during El Ni\~no years, including a reversal of rainfall anomalies between November of that year and January of the following one, influenced by land-surface interactions \citep{Grimm2003,Grimm2009}.
Even beyond El Ni\~no years, regional land-surface feedbacks can cause regions that exhibit wet anomalies in the spring to experience more summer precipitation on average \citep{Grimm2007}.
Similarly, mid-latitude dynamics influence low-level wind anomalies on many time scales, though this relationship is complicated due to coupled tropical-extratropical interactions \citep{Jones2002,Carvalho2004}.
To address these potential interactions, a cross-timescale approach based on synoptic circulation types is employed here to diagnose the causes of the rainfall events; this method has been used in previous work for southeastern South America \citep{Munoz2015,Munoz2016} and other regions \citep{Moron2015}.

We begin by describing our data sources in \cref{sec:data} and our methods in \cref{sec:methods}.
In \cref{sec:diagnostics} we start our diagnosis highlighting the observed flooding and contextualizing it within a long river stage time series; we then use composites and a weather typing analysis to diagnose the circulation patterns associated with the heavy rainfall during NDJF 2015-16.
We turn in \cref{sec:fcsts} to the question of whether the observed rainfall was successfully predicted by available models.
To carry out this analysis we study both forecasts targeting the entire series for a limited area, and also forecasts targeting a large spatial area for only the first week of December, when the most important flooding events started.
We also explore the impact on forecasts of several bias-correction schemes.
In \cref{sec:discussion} we discuss limitations and potential implications of our findings and potential future work, and we present our concluding remarks in \cref{sec:summary}.

@@

The analysis presented makes use of both observations and model forecasts.

@@

Rainfall data are taken from the CPC Unified Gauge-based Analysis of Global Daily Precipitation dataset \citep{Chen2008}.
Spatial resolution is \SI{0.5}{\degree} and temporal resolution is daily.
We define ``heavy'' rainfall events to be exceedances of the \nth{90} percentile; while the value is different for each grid cell, the \nth{90} percentile of area-averaged rainfall over the \gls{lprb} is approximately \SI{15}{\milli\meter\per\day}.

Atmospheric circulations are diagnosed using daily data from the NCAR-NCEP Reanalysis II dataset \citep{Kanamitsu2002}.
Spatial resolution is \SI{2.5}{\degree}.
Because the end-of-day time for the rainfall data is 12:00 GMT over most of South America \citep{Chen2008}, we use six-hour reanalysis data, and shift by twelve hours before re-sampling to the daily time step.
This ensures that the time steps in the reanalysis and ƒrainfall data sets are the same, but means that a day is defined as beginning at 12:00 GMT.
Since most summer rainfall in this region occurs overnight \citep{Vera2006,Salio2007}, this end-of-day time (which translates to approximately 8:00 AM locally depending on the exact time zone) tends to separate distinct events.
The primary atmospheric variable  used was the \SI{850}{\hecto\pascal} streamfunction, calculated directly from the wind field as described in \cref{sec:methods}.
The streamfunction is preferable to, for example, the geopotential height $\Phi$ because $\Phi$ has weak gradients near the equator, making it difficult to visualize circulations that span from the tropics to the extratropics.
The \SI{850}{\hecto\pascal} height level was used because it is representative of \gls{sallj} activity and moisture transport in this region \citep{Marengo2004,Salio2007}.

@@

This study also makes use of some climate indices.
Data on \gls{enso}, specifically the NINO3.4 index, came from a statistical-dynamical interpolation \citep{Kaplan1998}, which is constrained by relatively high-quality observations during the study period.
Data on the \gls{mjo}   came from the Australian Bureau of Meteorology \citep{Wheeler2004}.

@@

The seasonal predictions used  are known as ``flexible format'' forecasts, provided by the \gls{iri}.
These forecasts use a multi-model ensemble approach, with bias-corrected retrospective probabilistic forecasts produced using a total of (a) 144 members forced by evolving sea-surface temperatures, and (b) 68 members forced by persisted sea-surface temperatures; for details see \citet{Barnston2010}.
Flexible format means that the user of these forecasts can arbitrarily choose  particular thresholds (percentiles) to compute the  probability of exceedance (or non-exceedance) from the complete  probability density function of the climatological distribution,  rather than using the more common tercile categories.
The DJF 2015-2016 forecasts analyzed were produced in  November 2015.
Due to the short sample of flexible format forecasts available (only for 2012-2016 at the time of writing this paper), no verification was performed for these seasonal predictions.
These forecasts are provided at a horizontal  resolution of \SI{2.5}{\degree}.
The DJF 2015-2016 forecasts analyzed were produced in November 2015.

@@

Several types of analyses are used to diagnose the causes of the heavy rainfall events, and to bias-correct and verify the forecasts.
Computation was performed in the python environment using stable open source packages \citep{Hunter2007,McKinney2010,vanderWalt2011,Hoyer2017}.
All codes to reproduce or modify this analysis are available at \url{https://github.com/jdossgollin/PYFloods}.

@@

The $K$-means algorithm is guaranteed to converge to a local minimum of inter-cluster variance; to select the best partition, 500 simulations were created using the implementation in Python's \texttt{scikit-learn} package \citep{Pedregosa2012} and then the classifiability index of \citet{Michelangeli1995} was computed between each partition and the 499 others.
The partition whose classifiability index, averaged for all 499 pairwise comparisons, was the highest was selected.
Calculation of the classifiability index for several values of $K$ (supplemental figure S2) suggests that states with $K=5,6,\ldots,8$ are all reasonable.
We chose the solution $K=6$ because the clusters identified are qualitatively similar to those determined over southeastern South America \citep{Munoz2015,Munoz2016} and have an intuitive physical meaning, discussed further in the following sections.
We refer to the resulting clusters as \glspl{wt}.
From a physical point of view, the $K$-means algorithm helps identify typical atmospheric circulation patterns in the \gls{eof}-filtered field via clustering of days with similar streamfunction configurations.
These clusters can also be understood as proxies  of the available states of the system, or the most frequently visited trajectories in the phase space of the physical system \citep{Munoz2015,Munoz2016a,Munoz2017}.

@@

Logistic regression models the conditional probability of binary events, and it has been widely used in \gls{mos} approaches for binary outcomes.
Nonetheless, when using logistic regression to address multiple thresholds via independent fits, the predicted probabilities are, in general, not mutually consistent \citep{Messner2014}.
The \gls{xlr} was designed to address this shortcoming \citep{Wilks2009}, via the consideration of a transformation of the thresholds of interest as an additional predictor variable.
The \gls{hxlr}, a generalization of the \gls{xlr}, was proposed to appropriately use the ensemble spread as predictor for the dispersion of the predictive distribution \citep{Messner2014}.

@@

To evaluate model skill, we use a cross-validation approach with a 5-year window.
In this framework, five continuous years are left out of the record, the regression coefficients are computed with the remaining of the time series, and the resulting model is validated comparing the prediction for the third year left out (middle of the window) against observations.
The 5-year-long window is redefined a year at a time, moving from the beginning of the record to its end.

@@

\Cref{fig:streamflow} shows the streamflow time series at several gauges on the Paraguay River during NDJF 2015-16 in the context of their seasonality and decadal variability.
During November and December 2015, the river rose rapidly at Concepci\'on, Asunci\'on, and Pilar, though not at Bah\'ia Negra.
As discussed in \citet{Barros2004,Bravo2011}, the location of the Bah\'ia Negra gauge (see \cref{fig:study-area}) in the Pantanal region means that it responds very slowly to rainfall input.
The three downstream gauges, because they are located in the \gls{lprb}, respond to the rainfall forcing with a slow but steady rise.
Despite several very heavy storms, the streamflow record at Asunci\'on and Pilar (which are downstream of Concepci\'on) indicates relatively little response to individual storms.
Because the region is so flat (see topographic data in \cref{fig:study-area}), river levels at a particular point may be affected not only by rain in the catchment corresponding to that point, but also by elevated river levels downstream which reduce the pressure gradient available to drive flow.

@@

\Cref{fig:lagged-rain} shows time-lagged anomalies up to and after heavy rainfall dates (when area-averaged daily rainfall in the \gls{lprb} exceeds its NDJF \nth{90} percentile) and is consistent with previous analysis of heavy rainfall and intense convection in this region \citep{Liebmann2004,Marengo2004,Salio2007,Marwan2015}.
At $t=\SI{-2}{\day}$ a mid-latitude baroclinic system approaches the South American continent, intensifying and moving to the East from \SIrange{-1}{1}{\day}.
This system interacts with the sub-tropical low and the Andes Mountains to produce an anticyclonic anomaly over Brazil.
Along the cold front associated with this system, a low-level northerly jet advects  heat and moisture to the region.
As the system progresses, the jet below \SI{20}{\degree S} transitions from predominantly meridional flow (``Chaco Jet''; $t=\SI{-1}{\day}$ to predominantly zonal flow (``No-Chaco Jet'').
The pattern resembles composites identified using one standard deviation exceedances of rainfall at \SI{60}{\degree W}, \SI{30}{\degree S} \citep{Liebmann2004} and analysis for the \nth{95} or 99th percentiles of daily rainfall (not shown) yield similar results, implying that the synoptic mechanism for the most heavy events is not fundamentally distinct from the mechanism for moderate-intensity events.
This mean field, like all composites, masks between-event variation but exploration of individual events (not shown) indicates that the core features identified are generally present.

@@

\Gls{wt} 1 describes a \gls{sallj} event in which the strongest wind penetrates southward of \SI{25}{\degree S}, leading to heavy rainfall over NE Argentina and Uruguay; this has been called a ``Chaco Jet'' event \citep{Salio2002}.
\Gls{wt} 4 is also associated with a \gls{sallj} event, but the wind turns to the East northward of \SI{25}{\degree S}, leading to heavy rainfall over Eastern Paraguay and SW Brazil; this has been called a ``No-Chaco'' Jet event \citep{Vera2006}.
Supplemental Table S1 shows the centroids of each cluster, in the 4-dimensional phase space of the leading \glspl{eof} of \SI{850}{\hecto\pascal} streamfunction.

@@

We next use monthly-mean circulation anomalies (spatial patterns) and weather type sequences (temporal patterns) to understand the specific events of NDJF 2015-16.

@@

As indicated in \cref{fig:rain-wt}, the most heavy rainfall occurred during \glspl{wt} 1 and 4.
During NDJF 2015-16, \glspl{wt} 1 and 4 (Chaco and No-Chaco jet extensions, respectively),  occurred more frequently than their climatology (supplemental table S2); \gls{wt} 2 also occurred more frequently than its climatology, largely due to a long sequence in February 2016.
In mid-January 2016, during a sequence of persistent low rainfall, \gls{wt} 3 featured persistently, leading to heavy rainfall over central Brazil (not shown) and negative rainfall anomalies over the \gls{lprb}.
Thus while the intensity and persistence of heavy rainfall was atypical, the causal mechanism of the heavy rainfall observed during this season was consistent with climatology.

Inspection of \cref{fig:rain-wt} also suggests that at time scales of days to weeks, particular sequences of weather types tend to recur, and are associated with repeated rainfall storms.
The rainfall in late November 2015 occurred during \gls{wt} 1 and 4 days, with intermittent (and dry) \gls{wt} 5 occurrences.
During mid to late December 2015, \glspl{wt} 1 and 5 produced rainfall but the intermittent dry days were \gls{wt} 2.
During mid to late January 2016, \gls{wt} 3 led to persistent low rainfall.

@@

Heavy rainfall over the region was forecast for the DJF 2015-2016 season since at least November 2015 (see \cref{fig:seas-prob-fcst}).
Relative odds as high as 9:1 are visible over southern Paraguay and Brazil, and northern Uruguay and Argentina, broadly in agreement with observations.
The model predicted only very weakly increased odds of heavy rainfall in the Pantanal region (directly north of the \gls{lprb}) and in northern Argentina at $\approx \SI{65}{\degree W}$, and missed the heavy precipitation along most of the northeastern border of Paraguay.
However, the regionally elevated forecast of heavy rainfall could have been used for disaster preparedness at least one month in advance.

@@

\Cref{fig:chiclet} uses a Chiclet diagram \citep{Carbin2016} to visualize, as a function of lead time, the time evolution of the uncorrected, ensemble-mean rainfall anomaly forecast, spatially averaged over the \gls{lprb}.
At times greater than about two weeks, the ensemble-mean forecast is slightly wetter than climatology.
At weather timescales (less than one week), the ensemble-mean successfully predicts the timing and amplitude of the area-averaged rainfall.
At timescales of one to three weeks, the ensemble average successfully forecast the strongest breaks and pauses in the rainfall, such as the heavy rainfall during December 2015 and the dry period during mid-January 2016.
